### 2025-04-23 MSSC5750_Project.qmd
Adding initial file
importing data, loading a few packages, some basic data exploration

---
title: "Final Project"
author: "David, Michael, Patrick"
format: html
---

```{r echo = FALSE, message=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(scales)

# Importing data set from github repository
data <- read.csv("https://raw.githubusercontent.com/Pburns525/MSSC5750-Project/refs/heads/main/loan_data.csv")

```

High-Level Data Exploration
```{r}

nrow(data)
colnames(data)

# There is no missing data, already a cleaned data set
colSums(is.na(data))

table(data$person_gender)

# Education is not currently ordered, change to a factor and set levels
data$person_education<- factor(data$person_education, levels=c('High School','Associate','Bachelor','Master','Doctorate'), ordered=TRUE)
table(data$person_education)

# Looks like we have somebody who is 144 yrs old, 7 who are over age 100. PRobably want to remove them, its doubtful anybody very old would be applying anyways, and these ages are not really possible
summary(data$person_age)
sum(data$person_age>100)

summary(data$person_income)

summary(data$loan_amnt)

table(data$loan_intent)

summary(data$cb_person_cred_hist_length)

summary(data$loan_percent_income)

summary(data$credit_score)

table(data$person_home_ownership)

# This is years of employment experience. Again we see some super high values, longer than human lifespan, so should clean those up
summary(data$person_emp_exp)
sum(data$person_emp_exp>70)

table(data$previous_loan_defaults_on_file)

# 22% of our data receives a loan, so our dataset is imbalanced, but not too extreme
percent(table(data$loan_status)[2]/nrow(data),accuracy=1)

```

Creating some basic visualizations
```{R}

# Removing outliers (Somebody was earning like 7M, we have almost 25 who are over 1M still)
ggplot(data=data %>% filter(person_income<500000), aes(x=person_income)) +geom_histogram(bins=30)

#Actually might be better to just add a catch bucket at the end. IF over 250k, just set to 250k
ggplot(data=data %>% mutate(person_income2 = if_else(person_income<250000,person_income,250000)), aes(x=person_income2)) +geom_histogram(bins=40, fill='steelblue',color='black') + xlab("Income") + ggtitle("Histogram of Income")+
  scale_x_continuous(breaks=seq(0,250000,50000), labels=c("$0","$50,000","$100,000","$150,000","$200,000","$250,000+") )

```

### 2025-04-24 MSSC5750_Project.qmd
Add files via upload

@@
 library(scales)
+library(glmnet)
+library(fastDummies)
+library(caret)
 
@@
 ```
+
+ Outline for LASSO regression to help with variable selection
+```{R}
+
+# One-hot-encoding our categorical variables. Should happen automatically for factors
+data$person_gender <- as.factor(data$person_gender)
+data$person_home_ownership<-as.factor(data$person_home_ownership)
+data$loan_intent<-as.factor(data$loan_intent)
+data$previous_loan_defaults_on_file<-as.factor(data$previous_loan_defaults_on_file)
+
+# Extracting outcome variable
+class <- data$loan_status
+
+# Converting to a matrix
+matrixData <- model.matrix(class ~ ., data = data %>% select(-loan_status))
+
+# First column is all intercepts
+matrixData<- matrixData[,-1]
+
+# Uses cross validation, no need to train/test split. Alpha selects lasso (rather than ridge regression), but I think thats the default anyways
+lasso <- cv.glmnet(matrixData, class, alpha = 1, family = "binomial")
+
+# See coefficients, which variables we might drop
+coef(lasso)
+
+# Make predictions
+data$lassoPred <- as.vector(predict(lasso, newx = matrixData, type = "class"))
+
+# Evaluate accuracy measures
+confusionMatrix(factor(data$lassoPred), factor(data$loan_status), positive = '1')
+
+```

### 2025-04-24 MSSC5750_Project.qmd
Update MSSC5750_Project.qmd

@@
 library(caret)
+#
 
@@
 
-# One-hot-encoding our categorical variables. Should happen automatically for factors
+# Preparing categorical variables. Should happen automatically for factors
 data$person_gender <- as.factor(data$person_gender)

### 2025-04-29 MSSC5750_Project.qmd

--- 2025-04-24 MSSC5750_Project.qmd
@@
 ```
+
+Now doing a basic logistic regression with/without lasso selection
+```{R}
+
+# We need a train/test split. We will use 80/20
+# can use uniform distribution, set seed for reproducability
+set.seed(100)
+forTrain<-runif(nrow(data))<.8
+
+trainData<-data[forTrain,]
+testData<-data[!forTrain,]
+
+# First creating logistic regression with all our variables
+logModel <- glm(loan_status ~ person_age + person_gender + person_education + person_income + person_emp_exp + person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income + cb_person_cred_hist_length + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )
+
+# Repeating but removing the variables not relevant in lasso
+# removed(person cred length, person_emp_exp person_income person_gender person_education person_age)
+logModelLasso <- glm(loan_status ~ person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income  + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )
+
+testData$logisticPred <- round(predict(logModel, testData, type='response'))
+testData$logisticLassoPred <- round(predict(logModelLasso, testData, type='response'))
+
+# Matrix for regular log
+confusionMatrix(factor(testData$logisticPred), factor(testData$loan_status), positive = '1')
+
+# Matrix for lasso log
+confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1')
+
+```

### 2025-04-30 MSSC5750_Project.qmd
Added Shap Values & more Viz code

@@
 library(caret)
+library(shapr)
+library(shapviz)
+library(party)
+library(reshape2)
 
@@
 
+# Oddly higher credit score does not seem to make a difference
+cor(data$credit_score,data$loan_status)
+summary(data$credit_score)
+lowcred<-data %>% filter(credit_score<601)
+goodcred<-data %>% filter(credit_score>670)
+mean(lowcred$loan_status)
+mean(goodcred$loan_status)
+
 ```
@@
   scale_x_continuous(breaks=seq(0,250000,50000), labels=c("$0","$50,000","$100,000","$150,000","$200,000","$250,000+") )
+
+temp<-data %>% summarize('0'=1-mean(loan_status),"1"=mean(loan_status) ,.by=loan_intent)
+intent<-melt(temp) %>% dplyr::rename(loan_status=variable)
+
+ggplot(data=intent, aes(x=loan_intent, y=value, group=loan_status, fill=loan_status)) + geom_col(position='stack')
 
@@
 ```
+
+Shap values/viz
+```{r}
+# Requires split out xtrain, ytrain, xtest... Also a combined x/ytrain
+
+# We will apply to lasso data
+
+# We need to manually generate the dummy variables and match naems to what the model created
+# coder<-dummyVars(~ person_home_ownership + loan_intent + previous_loan_defaults_on_file, data = data, fullRank = TRUE, sep='')
+# trainDataAdj<-cbind(trainData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = trainData) )
+# testDataAdj<-cbind(testData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = testData) )
+
+# Only want the columns that we included in lasso
+
+xtrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)
+ytrain<-trainData %>% select(loan_status)
+xytrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file,loan_status)
+xtest<-testData  %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)
+
+# calculating shap is costly so we will reduce the size of the data we pass in
+
+# we will just explain on our train data as well for larger sample size
+set.seed(5)
+xtrainSmall <- xtrain[sample(nrow(xtrain), 1000), ]
+explainer<- shapr::explain(model=logModelLasso, x_train=xtrainSmall, x_explain=xtrainSmall, approach="ctree", phi0=mean(ytrain$loan_status))
+
+shp <- shapviz(explainer)
+sv_importance(shp)
+
+plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)
+
+shapObj <-shapviz(explainer)
+sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=15)
+
+```

### 2025-04-30 MSSC5750_Project.qmd
LDA

@@
 plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)
-
+####
 shapObj <-shapviz(explainer)

### 2025-05-01 MSSC5750_Project.qmd
Data Cleaning, One-Hot-Encoding, Results Table

--- 2025-04-30 MSSC5750_Project.qmd
@@
 
-table(data$previous_loan_defaults_on_file)
+# Hm. In every single case where there was a prior loan default, loan_status=0. We probably don't want to include this variable in anything explanatory
+table(data$previous_loan_defaults_on_file, data$loan_status)
 
@@
 
+Data cleaning
+```{R}
+
+data2<-data
+######## Transformation of variables ########
+
+#### person_income ####
+
+# Salary is extremely right skewed, lets use the log transformation. Minimum is 8000 so we do not need to add a constant
+min(data2$person_income)
+data2$log_person_income <- log(data2$person_income)
+
+# Our max is still way larger than our 3rd quartile, so maybe we should cap salary before transformation
+summary(data2$log_person_income)
+data2$log_person_income<-log(ifelse(data2$person_income>500000,500000,data2$person_income))
+
+# Better
+summary(data2$log_person_income)
+
+#### person_emp_exp ####
+
+# Employee experience also very skewed. And some of the values dont really make any sense. Log transformation can fix this too
+summary(data2$person_emp_exp)
+
+# First replace extreme (impossible) values with something realistic
+data2$person_emp_exp<-ifelse(data2$person_emp_exp>50,50,data2$person_emp_exp)
+
+data2$log_person_emp_exp <- log(data2$person_emp_exp+.5)
+
+# Better
+summary(data2$log_person_emp_exp)
+
+#### Age  ####
+# We need to bucket our age. Theres no way we actually have someone aged 144, and 75% of our data falls between 20-30
+data2<- data2 %>% mutate(person_age = case_when(person_age<25 ~ "<25",
+                                                person_age<=30 ~"25-30",
+                                                person_age<45 ~ "31-44",
+                                                person_age<60 ~ "45-59",
+                                                TRUE ~ "60+"))
+
+data2$person_age<-as.factor(data2$person_age)
+table(data2$person_age)
+
+####  cred hist length  ####
+# Only a handful fall >=18, we will just cap it all there as a "catch all", but leave as numeric
+table(data2$cb_person_cred_hist_length)
+data2$cb_person_cred_hist_length <- ifelse(data2$cb_person_cred_hist_length>18,18, data2$cb_person_cred_hist_length)
+
+######## One hot encoding for categorical variables ########
+
+# Preparing categorical variables.
+data2$person_gender <- as.factor(data2$person_gender)
+data2$person_home_ownership<-as.factor(data2$person_home_ownership)
+data2$loan_intent<-as.factor(data2$loan_intent)
+data2$previous_loan_defaults_on_file<-as.factor(data2$previous_loan_defaults_on_file)
+
+# One hot encoding for categorical variables
+data3 <- dummy_cols(data2, remove_first_dummy = FALSE, remove_selected_columns = TRUE)
+
+# we should remove first dummy to prevent multicolinnearity , need to keep track of what was dropped because that is our reference value
+data3<-data3 %>% select(-c("person_gender_female","person_education_High School","person_home_ownership_MORTGAGE", "loan_intent_DEBTCONSOLIDATION", "previous_loan_defaults_on_file_No", "person_age_<25"))
+
+# replacing base data with cleaner version now
+cleanData <- data3
+
+```
+
  Outline for LASSO regression to help with variable selection
@@
 
-# Preparing categorical variables. Should happen automatically for factors
-data$person_gender <- as.factor(data$person_gender)
-data$person_home_ownership<-as.factor(data$person_home_ownership)
-data$loan_intent<-as.factor(data$loan_intent)
-data$previous_loan_defaults_on_file<-as.factor(data$previous_loan_defaults_on_file)
+# Extracting outcome variable
+class <- cleanData$loan_status
 
-# Extracting outcome variable
-class <- data$loan_status
-
-# Converting to a matrix
-matrixData <- model.matrix(class ~ ., data = data %>% select(-loan_status))
+# Converting to a matrix, exclude the non-log-transformed, and also the prior loan default
+matrixData <- model.matrix(class ~ ., data = cleanData %>% select(-c("loan_status", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes") ))
 
@@
 # See coefficients, which variables we might drop
+# Now that we switched to log income, variable is no longer being dropped. We will proceed with log_income
 coef(lasso)
@@
 # Make predictions
-data$lassoPred <- as.vector(predict(lasso, newx = matrixData, type = "class"))
+cleanData$lassoPred <- as.vector(predict(lasso, newx = matrixData, type = "class"))
 
 # Evaluate accuracy measures
-confusionMatrix(factor(data$lassoPred), factor(data$loan_status), positive = '1')
+confusionMatrix(factor(cleanData$lassoPred), factor(cleanData$loan_status), positive = '1')
+
+# And again with our loan_default knowledge:
+confusionMatrix(factor( ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred)), factor(cleanData$loan_status), positive = '1')
 
@@
 set.seed(100)
-forTrain<-runif(nrow(data))<.8
+forTrain<-runif(nrow(cleanData))<.8
 
-trainData<-data[forTrain,]
-testData<-data[!forTrain,]
+trainData<-cleanData[forTrain,]
+testData<-cleanData[!forTrain,]
 
 # First creating logistic regression with all our variables
-logModel <- glm(loan_status ~ person_age + person_gender + person_education + person_income + person_emp_exp + person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income + cb_person_cred_hist_length + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )
+logModel <- glm(loan_status ~ . , data=trainData %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes")), family= binomial )
 
-# Repeating but removing the variables not relevant in lasso
-# removed(person cred length, person_emp_exp person_income person_gender person_education person_age)
-logModelLasso <- glm(loan_status ~ person_home_ownership +loan_amnt + loan_intent + loan_int_rate + loan_percent_income  + credit_score + previous_loan_defaults_on_file, data=trainData, family= binomial )
+# Repeating but only with variables selected by lasso
+logModelLasso <- glm(loan_status ~ loan_int_rate + loan_percent_income + log_person_income + person_home_ownership_OTHER + person_home_ownership_OWN + person_home_ownership_RENT + loan_intent_EDUCATION + loan_intent_HOMEIMPROVEMENT + loan_intent_PERSONAL + loan_intent_VENTURE, data=trainData, family= binomial )
 
@@
 confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1')
+
+# Our training data is still heavily imbalanced, repeating again for rebalanced data
+table(trainData$loan_status)
+ones<-trainData %>% filter(loan_status==1)
+set.seed(5)
+zeros<-trainData %>% filter(loan_status==0) %>% sample_n(8000)
+trainDataBalanced<-rbind(ones,zeros)
+table(trainDataBalanced$loan_status)
+
+rebalancedLog <- glm(loan_status ~ . , data=trainDataBalanced %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes")), family= binomial )
+
+testData$rebalancedLog <- round(predict(rebalancedLog, testData, type='response'))
+
+# Matrix for regular log
+confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1')
+
+```
+
+Creating dataframe of results
+```{R}
+
+results<-data.frame(model=character(),accuracy=numeric(),sensitivity=numeric(),specificity=numeric())
+
+addresults<-function(name,res){
+  results<<-rbind(results, data.frame(model=name,accuracy=res$overall['Accuracy'],sensitivity=res$byClass['Sensitivity'],specificity=res$byClass['Specificity']) )
+}
+
+addresults('Lasso', confusionMatrix(factor( cleanData$lassoPred), factor(cleanData$loan_status), positive = '1'))
+addresults('Log', confusionMatrix(factor(testData$logisticPred), factor(testData$loan_status), positive = '1'))
+addresults('LogLasso', confusionMatrix(factor(testData$logisticLassoPred), factor(testData$loan_status), positive = '1'))
+addresults('LogBalanced', confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1'))
+rownames(results)<-NULL
+results
+
+###################### Repeating for accuracy after applying logic for previous default ##############################
+
+results2<-data.frame(model=character(),accuracy=numeric(),sensitivity=numeric(),specificity=numeric())
+
+addresults2<-function(name,res){
+  results2<<-rbind(results2, data.frame(model=name,accuracy=res$overall['Accuracy'],sensitivity=res$byClass['Sensitivity'],specificity=res$byClass['Specificity']) )
+}
+
+addresults2('Lasso', confusionMatrix(factor(ifelse(cleanData$previous_loan_defaults_on_file_Yes==1,0,cleanData$lassoPred) ), factor(cleanData$loan_status), positive = '1'))
+addresults2('Log', confusionMatrix(factor( ifelse(testData$previous_loan_defaults_on_file_Yes==1,0, testData$logisticPred) ), factor(testData$loan_status), positive = '1'))
+addresults2('LogLasso', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$logisticLassoPred) ), factor(testData$loan_status), positive = '1'))
+addresults2('LogBalanced', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$rebalancedLog) ), factor(testData$loan_status), positive = '1'))
+rownames(results2)<-NULL
+results2
 
@@
 plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)
+####
 
+###
 shapObj <-shapviz(explainer)

### 2025-05-02 MSSC5750_Project.qmd
Fixing Shap Values

@@
 ```{r}
-# Requires split out xtrain, ytrain, xtest... Also a combined x/ytrain
+# Requires split out xtrain, ytrain, xtest.
 
-# We will apply to lasso data
+xtrain<-trainDataBalanced %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes", "loan_status"))
+ytrain<-trainDataBalanced %>% select(loan_status)
 
-# We need to manually generate the dummy variables and match naems to what the model created
-# coder<-dummyVars(~ person_home_ownership + loan_intent + previous_loan_defaults_on_file, data = data, fullRank = TRUE, sep='')
-# trainDataAdj<-cbind(trainData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = trainData) )
-# testDataAdj<-cbind(testData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = testData) )
-
-# Only want the columns that we included in lasso
-
-xtrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)
-ytrain<-trainData %>% select(loan_status)
-xytrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file,loan_status)
-xtest<-testData  %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)
+xtest<-testData  %>% select(-c("lassoPred", "person_income", "person_emp_exp", "previous_loan_defaults_on_file_Yes", "loan_status", "logisticPred","logisticLassoPred", "rebalancedLog"))
 
@@
 set.seed(5)
-xtrainSmall <- xtrain[sample(nrow(xtrain), 1000), ]
-explainer<- shapr::explain(model=logModelLasso, x_train=xtrainSmall, x_explain=xtrainSmall, approach="ctree", phi0=mean(ytrain$loan_status))
+xtestSmall <- xtest[sample(nrow(xtest), 250), ]
+explainer<- shapr::explain(model=rebalancedLog, x_train=xtrain, x_explain=xtestSmall, approach="ctree", phi0=mean(ytrain$loan_status))
 
@@
 shapObj <-shapviz(explainer)
-sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=15)
+sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=25)
+sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=10)
 

### 2025-05-03 MSSC5750_Project.qmd
More Visualizations/Transformations

@@
 library(reshape2)
+library(gridExtra)
 
@@
 
+# Looking at age/credit score with loan approval. NO clear patterns
+ggplot(data=rbind( data %>% filter(loan_status==1) %>%  sample_n(1000), data %>% filter(loan_status==0) %>%  sample_n(1000)), aes(x=credit_score, y=person_age, fill=loan_status, color=loan_status)) + 
+  geom_point() + ylim(20,60) + xlim(400,780)
+
 ```
@@
 
-#Actually might be better to just add a catch bucket at the end. IF over 250k, just set to 250k
-ggplot(data=data %>% mutate(person_income2 = if_else(person_income<250000,person_income,250000)), aes(x=person_income2)) +geom_histogram(bins=40, fill='steelblue',color='black') + xlab("Income") + ggtitle("Histogram of Income")+
-  scale_x_continuous(breaks=seq(0,250000,50000), labels=c("$0","$50,000","$100,000","$150,000","$200,000","$250,000+") )
+#Actually might be better to just add a catch bucket at the end. IF over 500k, just set to 500k
+inc1<-ggplot(data=data %>% mutate(person_income2 = if_else(person_income<500000,person_income,500000)), aes(x=person_income2)) +geom_histogram(bins=40, fill='steelblue',color='black') + xlab("Income") + ggtitle("Histogram of Income") +
+  scale_x_continuous(breaks=seq(0,500000,50000), labels=c("$0","$50,000","$100,000","$150,000","$200,000","$250,000","$300,000","$350,000","$400,000","$450,000", "$500,000+") )
 
+# Visual for log adjusted income
+inc2<-ggplot(data = data %>% mutate('log(person_income)' = log(ifelse(person_income>500000,500000,person_income))  ), aes(x=log(person_income))) +geom_histogram(bins=40, fill='orange',color='black') + xlab("Log of Income") + ggtitle("Histogram of Log of Income")
+
+grid.arrange(inc1, inc2, nrow = 2)
+
+# Loan intent
 temp<-data %>% summarize('0'=1-mean(loan_status),"1"=mean(loan_status) ,.by=loan_intent)
@@
 
-ggplot(data=intent, aes(x=loan_intent, y=value, group=loan_status, fill=loan_status)) + geom_col(position='stack')
+ggplot(data=intent %>% filter(loan_status==1), aes(x=loan_intent, y=value)) + 
+  geom_col(position='stack', fill="violetred2") + 
+  geom_text(aes(label = percent(value,accuracy=1)), 
+            position = position_stack(vjust = 0.5), 
+            size = 4) + ggtitle('Loan Repayment by Intent') + 
+  scale_y_continuous(name='', labels = scales::percent, limits=c(0,.4)) + xlab('') +
+  theme(panel.grid.minor.y=element_blank())
+
+# Education
+temp<-data %>% summarize('0'=1-mean(loan_status),"1"=mean(loan_status) ,.by=person_education)
+educ<-melt(temp) %>% dplyr::rename(loan_status=variable)
+
+ggplot(data=educ %>% filter(loan_status==1), aes(x=person_education, y=value)) +
+    geom_col(position='stack', fill='aquamarine3') +
+    geom_text(aes(label = percent(value,accuracy=1)), 
+            position = position_stack(vjust = 0.5), 
+            size = 4) + ggtitle('Loan Repayment by Education') + 
+  scale_y_continuous(name='', labels = scales::percent, limits=c(0,.4)) + xlab('') +
+  theme(panel.grid.minor.y=element_blank())
+
+### cred hist length
+ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue')
 

### 2025-05-03 MSSC5750_Project.qmd
Update MSSC5750_Project.qmd

@@
 ```
+
+## ADD CODE FOR LDA
+
+library(tidyverse)
+library(MASS)
+bankr<-data.frame(loan_data$person_age,loan_data$person_income,loan_data$person_emp_exp,loan_data$loan_amnt,
+                  loan_data$loan_int_rate,loan_data$loan_percent_income,loan_data$cb_person_cred_hist_length,
+                  loan_data$credit_score,loan_data$previous_loan_defaults_on_file)
+
+head(bankr)
+
+lda_desc<-lda(loan_data$previous_loan_defaults_on_file~loan_data$person_age+
+                loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+                loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+                loan_data$credit_score,data=bankr )
+
+lda_desc
+
+a<-lda_desc$scaling
+a
+
+library(MVN)
+library(tidyverse)
+
+previous_default<-bankr%>%
+  filter(loan_data.previous_loan_defaults_on_file=="Yes")%>%
+  select(-loan_data.previous_loan_defaults_on_file)
+  s1<-cov(previous_default)
+
+  s1
+
+previous_default
+# Checking for multivariate normality
+defaulterscheck<-mvn(previous_default[,1:8],mvnTest = "mardia")$multivariateNormality
+print(defaulterscheck)
+
+#separate groups into two and compute sample covariance matrix
+no_default<-bankr%>%
+  filter(loan_data.previous_loan_defaults_on_file =="No")%>%
+  select(-loan_data.previous_loan_defaults_on_file)
+s2<-cov(no_default)
+s2
+
+# pooled sample covariance matrix  is :
+
+n1<-nrow(previous_default)
+
+n2<-nrow(no_default)
+
+Spl<-1/(n1+n2-2)*((n1-1)*s1 +(n2-1)*s2)
+
+Spl
+
+# manually calculating the coefficients of the standardized discriminant function
+
+a.star<-sqrt(diag(Spl))*a
+
+a.star
+# a.star ranks the independependent variables in the order of discriminative capacity
+# i.e which variables best separate the groups: load_default on file vs no history of default
+
+# 1 age
+#2. income
+#3. employment experience
+#4. loan amount
+#5. loan interest rate
+# 6. loan percent of income (i.e debt-to-income ratio)
+#7.  length of credit history
+# 8. Credit score
+
+library(tidyverse)
+ggplot(bankr,aes(x=loan_data$person_age,y=loan_data$loan_percent_income,
+                 color=factor(loan_data$previous_loan_defaults_on_file)))+
+  geom_point()+ labs(xlab="Age",ylab = "debt-to-incomes")+
+  scale_color_discrete(name="loan_data$previous_loan_defaults_on_file")
+
+non_defaultcheck<-mvn(no_default[,1:8],mvnTest = "mardia")$multivariateNormality
+
+print(non_defaultcheck)
+
+pca_bank<-prcomp(bankr[,1:8],scale. = TRUE)
+summary(pca_bank)
+
+pca_bank # pca to confirm the variability in data set and ranking of variables 
+           # in discrimination  b/w 2 groups
+
+lda2<-lda(loan_data.previous_loan_defaults_on_file~loan_data.person_age + loan_data.person_income+
+            loan_data.person_emp_exp +loan_data.loan_amnt,data =bankr)
+
+lda2
+
+library(MASS)
+library(tidyverse)
+library(Hotelling)
+
+loans.test<-hotelling.test(loan_data$person_age+
+    loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+    loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+    loan_data$credit_score~factor(loan_data$previous_loan_defaults_on_file),data=bankr )
+
+loans.test
+
+#LDA-Predict groups-taking a predictive view
+
+lda3<-lda(loan_data$previous_loan_defaults_on_file~loan_data$person_age+
+            loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+            loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+            loan_data$credit_score,prior=c(0.5,0.5),data=bankr )
+lda3.results<-bankr%>%
+  mutate(classify.loan_data.previous_loan_defaults_on_file=predict(lda3)$class)
+
+lda3.results
+
+lda3$scaling
+
+zbar1 <- sum(lda3$scaling*lda3$means[1,])
+zbar1
+
+zbar2<- sum(lda3$scaling*lda3$means[2,])
+
+zbar2
+
+cutoff<-0.5*(zbar1+zbar2)
+
+cutoff
+
+#checking classification results
+with(lda3.results,table(loan_data.previous_loan_defaults_on_file,classify.loan_data.previous_loan_defaults_on_file))
+
+#Classification Analysis Result
+
+# overall classification  (13,508 + 14793)/45,000 : 63% Correct classification
+
+# No default :13508/(13,508+8,634)
+
+# Previous default : 14,793/(14,793+8,065)
+
+pairs(bankr[,1:8])

### 2025-05-03 MSSC5750_Project.qmd
David Code Variable Swap

--- 2025-05-03 MSSC5750_Project.qmd
@@
 ### cred hist length
-ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue')
+ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue') + ggtitle('Frequencies for credit history length')
 
@@
 addresults('LogBalanced', confusionMatrix(factor(testData$rebalancedLog), factor(testData$loan_status), positive = '1'))
-rownames(results)<-NULL
 results
@@
 addresults2('LogBalanced', confusionMatrix(factor(ifelse(testData$previous_loan_defaults_on_file_Yes==1,0,testData$rebalancedLog) ), factor(testData$loan_status), positive = '1'))
-rownames(results2)<-NULL
 results2
@@
 ```
+
+## ADD CODE FOR LDA
+
+```{r}
+
+library(tidyverse)
+library(MASS)
+loan_data <- read.csv("https://raw.githubusercontent.com/Pburns525/MSSC5750-Project/refs/heads/main/loan_data.csv")
+bankr<-data.frame(loan_data$person_age,loan_data$person_income,loan_data$person_emp_exp,loan_data$loan_amnt,
+                  loan_data$loan_int_rate,loan_data$loan_percent_income,loan_data$cb_person_cred_hist_length,
+                  loan_data$credit_score,loan_data$loan_status)
+
+head(bankr)
+
+lda_desc<-lda(loan_data$loan_status~loan_data$person_age+
+                loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+                loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+                loan_data$credit_score,data=bankr )
+
+lda_desc
+
+a<-lda_desc$scaling
+a
+
+library(MVN)
+library(tidyverse)
+
+# renaming previous_default to loan_approved
+# previous name previous_default
+loan_repaid<-bankr %>%
+  filter(loan_data.loan_status==1) %>%
+  select(-loan_data.loan_status)
+  s1<-cov(loan_repaid)
+
+  s1
+
+loan_repaid
+# Checking for multivariate normality
+repaycheck<-mvn(loan_repaid[,1:8],mvnTest = "mardia")$multivariateNormality
+print(repaycheck)
+
+#separate groups into two and compute sample covariance matrix
+# previous name no_default
+loan_default<-bankr%>%
+  filter(loan_data.loan_status ==0)%>%
+  select(-loan_data.loan_status)
+s2<-cov(loan_default)
+s2
+
+# pooled sample covariance matrix  is :
+
+n1<-nrow(loan_repaid)
+
+n2<-nrow(loan_default)
+
+Spl<-1/(n1+n2-2)*((n1-1)*s1 +(n2-1)*s2)
+
+Spl
+
+# manually calculating the coefficients of the standardized discriminant function
+
+a.star<-sqrt(diag(Spl))*a
+
+a.star
+# a.star ranks the independependent variables in the order of discriminative capacity
+# i.e which variables best separate the groups: load_default on file vs no history of default
+
+# 1 age
+#2. income
+#3. employment experience
+#4. loan amount
+#5. loan interest rate
+# 6. loan percent of income (i.e debt-to-income ratio)
+#7.  length of credit history
+# 8. Credit score
+
+library(tidyverse)
+ggplot(bankr,aes(x=loan_data$person_age,y=loan_data$loan_percent_income,
+                 color=factor(loan_data$loan_status)))+
+  geom_point()+ labs(xlab="Age",ylab = "debt-to-incomes")+
+  scale_color_discrete(name="loan_data$loan_status")
+
+repaidCheck<-mvn(loan_repaid[,1:8],mvnTest = "mardia")$multivariateNormality
+
+print(repaidCheck)
+
+pca_bank<-prcomp(bankr[,1:8],scale. = TRUE)
+summary(pca_bank)
+
+pca_bank # pca to confirm the variability in data set and ranking of variables 
+           # in discrimination  b/w 2 groups
+
+lda2<-lda(loan_data.loan_status~loan_data.person_age + loan_data.person_income+
+            loan_data.person_emp_exp +loan_data.loan_amnt,data =bankr)
+
+lda2
+
+library(MASS)
+library(tidyverse)
+library(Hotelling)
+
+loans.test<-hotelling.test(loan_data$person_age+
+    loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+    loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+    loan_data$credit_score~factor(loan_data$loan_status),data=bankr )
+
+loans.test
+
+#LDA-Predict groups-taking a predictive view
+
+lda3<-lda(loan_data$loan_status~loan_data$person_age+
+            loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
+            loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
+            loan_data$credit_score,prior=c(0.5,0.5),data=bankr )
+lda3.results<-bankr%>%
+  mutate(classify.loan_data.loan_status=predict(lda3)$class)
+
+lda3.results
+
+lda3$scaling
+
+zbar1 <- sum(lda3$scaling*lda3$means[1,])
+zbar1
+
+zbar2<- sum(lda3$scaling*lda3$means[2,])
+
+zbar2
+
+cutoff<-0.5*(zbar1+zbar2)
+
+cutoff
+
+#checking classification results
+with(lda3.results,table(loan_data.loan_status,classify.loan_data.loan_status))
+
+#Classification Analysis Result
+
+# overall classification  (13,508 + 14793)/45,000 : 63% Correct classification
+
+# No default :13508/(13,508+8,634)
+
+# Previous default : 14,793/(14,793+8,065)
+
+pairs(bankr[,1:8])
+
+```
+
+Adding LDA results to table
+```{R}
+
+addresults('LDA', confusionMatrix(factor( lda3.results$classify.loan_data.loan_status), factor(lda3.results$loan_data.loan_status), positive = '1'))
+rownames(results)<-NULL
+
+# rejoining to get back the previous default data
+temp <- lda3.results %>% left_join(data %>% select(person_age, person_income, loan_percent_income, person_emp_exp,credit_score,loan_int_rate,previous_loan_defaults_on_file), by=c("loan_data.person_age"="person_age", "loan_data.person_income"="person_income", "loan_data.loan_percent_income"="loan_percent_income", "loan_data.person_emp_exp" = "person_emp_exp", "loan_data.credit_score"="credit_score", "loan_data.loan_int_rate"= "loan_int_rate")) %>% mutate(classify.loan_data.loan_status=as.numeric(classify.loan_data.loan_status)-1)
+
+addresults2('LDA', confusionMatrix(factor( ifelse(temp$previous_loan_defaults_on_file=='Yes',0,temp$classify.loan_data.loan_status ) ), factor(temp$loan_data.loan_status), positive = '1'))
+rownames(results2)<-NULL
+```

### 2025-05-05 MSSC5750_Project.qmd
Final Updates and Cleanup

@@
 
+### quick summarizations to give us an idea of what we need to visualize/look closer at
 summary(data$person_income)
@@
 ### cred hist length
-ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue') + ggtitle('Frequencies for credit history length')
+ggplot(data=data %>% summarize(count=n(),.by=cb_person_cred_hist_length), aes(x= cb_person_cred_hist_length, y=count)) + geom_col(fill='steelblue') + ggtitle('Frequencies for credit history length') + scale_x_continuous(breaks=c(2,seq(10,30,5)) )
 
@@
 ```{r}
+# Originally was using prior loan default as outcome variable
+# Switched to loan_status to match the target variable we used for everything else
 
@@
 
-Adding LDA results to table
+Adding David's LDA results to table
 ```{R}
@@
 rownames(results2)<-NULL
+
+cbind("Model"=results[,1],round(results[,2:4],3))
+cbind("Model"=results2[,1],round(results2[,2:4],3))
 ```
